\documentclass[12pt,a4paper]{article}
\usepackage[utf8]{inputenc}
\usepackage[T1]{fontenc}
\usepackage{amsmath,amssymb,amsthm}
\usepackage{geometry}
\usepackage{enumitem}
\usepackage{hyperref}
\usepackage{fancyhdr}
\usepackage{titlesec}

\geometry{margin=1in}
\pagestyle{fancy}
\fancyhf{}
\rhead{Lecture 4}
\lhead{Quantum Systems via Linear Algebra}
\cfoot{\thepage}

\titleformat{\section}{\large\bfseries}{}{0em}{}
\titleformat{\subsection}{\normalsize\bfseries}{}{0em}{}

\newtheorem{definition}{Definition}
\newtheorem{theorem}{Theorem}

\title{\textbf{Lecture 4: Time Evolution as an LTI System} \\ \large The System Matrix, Unitary Evolution, and the Schr\"{}dinger ODE}
\author{Quantum Systems via Linear Algebra, Differential Equations, and Probability}
\date{}

\begin{document}
\maketitle
\thispagestyle{fancy}

% ============================================================
\section{1. The Evolution ODE}
% ============================================================

Axiom~4 (Lecture~1) states that an isolated system evolves by
\[
    i\hbar\,\dot{\mathbf{v}}(t) = H\,\mathbf{v}(t),
\]
where $H = H^\dagger$ is the system matrix (coefficient matrix of the evolution ODE).  This is a first-order linear system with constant coefficients---engineers solve these routinely.

\subsection{1.1 Solution via matrix exponential}
\[
    \mathbf{v}(t) = e^{-iHt/\hbar}\,\mathbf{v}(0).
\]
Call $U(t) = e^{-iHt/\hbar}$ the \textbf{evolution operator}.

\subsection{1.2 Unitarity and norm preservation}

\begin{theorem}
$U(t) = e^{-iHt/\hbar}$ is unitary ($U^\dagger U = I$) if and only if $H = H^\dagger$.
\end{theorem}

\noindent\textbf{Proof.}  For an infinitesimal step $\epsilon$: $U(\epsilon) = I - i\epsilon H/\hbar + \mathcal{O}(\epsilon^2)$.  Then $U^\dagger(\epsilon) = I + i\epsilon H^\dagger/\hbar + \mathcal{O}(\epsilon^2)$.  Imposing $U^\dagger U = I$ to first order:
\[
    I + \frac{i\epsilon}{\hbar}(H^\dagger - H) = I \quad\Longrightarrow\quad H^\dagger = H.\;\square
\]

\noindent\textbf{The physics direction:} We \emph{demand} unitarity because probability must be conserved---if $\|\mathbf{v}(0)\|=1$, then $\|\mathbf{v}(t)\|=1$ for all $t$.  Self-adjointness of $H$ is the \emph{consequence}, not the assumption.  This completes the chain promised in Lecture~1: norm preservation $\Rightarrow$ unitarity $\Rightarrow$ Hermitian generator.

\subsection{1.3 Important caveat}
This describes evolution of an \emph{isolated} system---one not being measured or interacting with anything external.  During a measurement, the state update rule (Lecture~1, Section~3.1) applies instead.

% ============================================================
\section{2. Deriving the ODE from Unitarity}
% ============================================================

Starting from
\[
    \mathbf{v}(\epsilon) = (I - i\epsilon H/\hbar)\,\mathbf{v}(0),
\]
rearrange:
\[
    \frac{\mathbf{v}(\epsilon)-\mathbf{v}(0)}{\epsilon} = -\frac{i}{\hbar}H\,\mathbf{v}(0).
\]
Taking $\epsilon\to 0$:
\[
    \boxed{i\hbar\,\frac{d}{dt}\mathbf{v}(t) = H\,\mathbf{v}(t).}
\]
This is the \textbf{time-dependent Schr\"{o}dinger equation}---a first-order linear ODE with constant coefficient matrix $H$.

\medskip
\noindent\textbf{Engineering connection:} This has exactly the form of a state-space model $\dot{\mathbf{x}} = A\mathbf{x}$ with $A = -iH/\hbar$.  The solution $\mathbf{v}(t) = e^{At}\mathbf{v}(0)$ is the standard matrix exponential.  The difference from a typical LTI system is that $A$ is skew-Hermitian ($A^\dagger = -A$), so the evolution is norm-preserving rather than dissipative.

% ============================================================
\section{3. Energy Eigenstates and the General Solution}
% ============================================================

The system matrix $H$ is a self-adjoint matrix representing energy.  Its eigenvalue equation
\[
    \boxed{H\mathbf{e}_k = E_k\,\mathbf{e}_k}
\]
gives the allowed energy levels $\{E_k\}$ and the energy eigenvectors $\{\mathbf{e}_k\}$ (a complete orthonormal basis).

\subsection{3.1 Stationary states}
If $\mathbf{v}(0) = \mathbf{e}_k$, then
\[
    \mathbf{v}(t) = e^{-iE_k t/\hbar}\,\mathbf{e}_k.
\]
This is a phase rotation---no change in measurement probabilities.  Energy eigenstates are \textbf{stationary}: all observables have time-independent statistics.

\medskip
\noindent\textbf{Engineering analog:} Energy eigenstates are the \emph{normal modes} of the system.  Each eigenvalue $E_k$ is an energy, and it determines the oscillation frequency of that mode: $\omega_k = E_k/\hbar$ (or simply $\omega_k = E_k$ with $\hbar=1$).  This is the quantum version of $E = \hbar\omega$---the eigenvalues of $H$ are energies, and their ratios to $\hbar$ are the natural frequencies (poles) of the system.

\subsection{3.2 General solution}
Expand the initial state: $\mathbf{v}(0) = \sum_k \alpha_k\,\mathbf{e}_k$.  Then
\[
    \boxed{\mathbf{v}(t) = \sum_k \alpha_k\,e^{-iE_k t/\hbar}\,\mathbf{e}_k.}
\]

\medskip
\noindent\textbf{The central insight of unitary evolution.}  The magnitudes $|\alpha_k|$ do not change---all time dependence is in the relative phases $e^{-iE_k t/\hbar}$.  In the energy eigenbasis, time evolution is \emph{trivial}: it is $N$ independent phase rotations at $N$ different frequencies $\omega_k = E_k/\hbar$.  Every quantum dynamical phenomenon---Rabi oscillations, interference, beats, revivals---is a consequence of those phase rotations running at \emph{different rates} while the observable being measured has eigenvectors that are \emph{misaligned} with the energy eigenvectors.

If you measure an observable that commutes with $H$, you see no dynamics: the energy eigenbasis is also the eigenbasis of your observable, and the relative phases cancel in $|\mathbf{e}_i^\dagger\mathbf{v}|^2$.  All quantum dynamics comes from measuring observables whose eigenbasis is \emph{misaligned} with the energy eigenbasis.  Misalignment plus phase rotation equals interference equals dynamics.

\medskip
\noindent\fbox{\parbox{\dimexpr\linewidth-2\fboxsep-2\fboxrule}{%
\textbf{Code says:} Computing $e^{-iHt/\hbar}$ numerically: diagonalize $H = V\Lambda V^\dagger$, then $e^{-iHt/\hbar} = V\,e^{-i\Lambda t/\hbar}\,V^\dagger$ where $e^{-i\Lambda t/\hbar}$ is diagonal with entries $e^{-i\lambda_k t/\hbar}$.  In Python: \texttt{scipy.linalg.expm(-1j * H * t / hbar)}.}}

\subsection{3.3 Worked example: Rabi oscillations}
Set $H = \frac{\omega}{2}X$ (with $\hbar=1$ for simplicity).  Take initial state $\mathbf{v}(0)=\mathbf{e}_0$.  The eigenvectors of $X$ are $\mathbf{x}_\pm = \frac{1}{\sqrt{2}}(\mathbf{e}_0\pm\mathbf{e}_1)$ with eigenvalues $\pm 1$, so the energy eigenvalues are $E_\pm = \pm\omega/2$.

\noindent\textbf{Step 1: expand in energy eigenbasis.}
\[
    \mathbf{e}_0 = \tfrac{1}{\sqrt{2}}(\mathbf{x}_++\mathbf{x}_-).
\]
\textbf{Step 2: apply phase rotations.}
\[
    \mathbf{v}(t) = \tfrac{1}{\sqrt{2}}\bigl(e^{-i\omega t/2}\,\mathbf{x}_+ + e^{+i\omega t/2}\,\mathbf{x}_-\bigr).
\]
\textbf{Step 3: compute $\langle Z\rangle(t) = \mathbf{v}(t)^\dagger Z\,\mathbf{v}(t)$.}  Substituting and using $\mathbf{e}_0^\dagger Z\mathbf{e}_0=1$, $\mathbf{e}_1^\dagger Z\mathbf{e}_1=-1$:
\[
    \langle Z\rangle(t) = |\alpha_0(t)|^2 - |\alpha_1(t)|^2 = \cos(\omega t).
\]
This is the simplest \textbf{Rabi oscillation}---the system oscillates between $\mathbf{e}_0$ and $\mathbf{e}_1$ at frequency $\omega$.  The dynamics arise because $Z$ and $H=\frac{\omega}{2}X$ do not share an eigenbasis ($[Z,X]\neq 0$).

\medskip
\noindent\fbox{\parbox{\dimexpr\linewidth-2\fboxsep-2\fboxrule}{%
\textbf{Code says:} Twelve lines of Python produce the $\cos(\omega t)$ curve---the reader's first simulation of a quantum system.  See the companion notebook.}}

% ============================================================
\section{4. Inner Product Preservation}
% ============================================================

If $\mathbf{u}^\dagger\mathbf{w} = 0$ at $t=0$ (states initially distinguishable), then $\mathbf{u}(t)^\dagger\mathbf{w}(t) = 0$ for all $t$.  More generally:
\[
    \mathbf{u}(t)^\dagger\mathbf{w}(t) = \mathbf{u}(0)^\dagger\,U^\dagger(t)\,U(t)\,\mathbf{w}(0) = \mathbf{u}(0)^\dagger\mathbf{w}(0).
\]
The ``overlap'' between any two states is preserved.  Unitary evolution cannot create or destroy \textbf{distinguishability}: two states that could be perfectly told apart at $t=0$ remain perfectly distinguishable forever, and two states that were indistinguishable remain so.  This is the precise sense in which quantum evolution is \textbf{information-preserving}---the foundation of quantum information theory, the reason the no-cloning theorem holds, and the reason Hawking's information paradox is a paradox.

\medskip
\noindent\textbf{Engineering analog:} Unitary evolution $U^\dagger U = I$ is analogous to a lossless network or power-preserving two-port ($S^\dagger S = I$ in microwave engineering).  Both preserve a quadratic form; both are reversible.

% ============================================================
\section{5. Expectation Values and Their Dynamics}
% ============================================================

\subsection{5.1 Expectation value as a quadratic form}
The average value of observable $L$ in state $\mathbf{v}$ is the Rayleigh quotient:
\[
    \boxed{\langle L \rangle = \mathbf{v}^\dagger L\,\mathbf{v} = \sum_i \lambda_i\,|\mathbf{e}_i^\dagger\mathbf{v}|^2.}
\]
\textbf{Proof:} Expand $\mathbf{v}=\sum_i \alpha_i\mathbf{e}_i$ in eigenvectors of $L$.  Then $\mathbf{v}^\dagger L\mathbf{v} = \sum_{i,j}\alpha_j^*\alpha_i\lambda_i\,\mathbf{e}_j^\dagger\mathbf{e}_i = \sum_i |\alpha_i|^2\lambda_i$.\;$\square$

\subsection{5.2 Time evolution of expectation values}
Let $\langle L \rangle(t) = \mathbf{v}(t)^\dagger L\,\mathbf{v}(t)$.  Differentiating using $\dot{\mathbf{v}}=-\frac{i}{\hbar}H\mathbf{v}$ and $\dot{\mathbf{v}}^\dagger=+\frac{i}{\hbar}\mathbf{v}^\dagger H$:
\[
    \frac{d}{dt}\langle L \rangle = \frac{i}{\hbar}\mathbf{v}^\dagger HL\mathbf{v} - \frac{i}{\hbar}\mathbf{v}^\dagger LH\mathbf{v} = \frac{i}{\hbar}\langle [H,L] \rangle,
\]
where $[H,L]\equiv HL - LH$ is the \textbf{commutator}.  So:
\[
    \boxed{\frac{d}{dt}\langle L \rangle = \frac{i}{\hbar}\langle [H,L] \rangle.}
\]
This is the \textbf{Ehrenfest equation}---the bridge between quantum evolution and classical equations of motion.

\subsection{5.3 Worked example: Bloch precession}
Set $H = \frac{\Omega}{2}Z$ (with $\hbar=1$).  Compute $\frac{d}{dt}\langle X\rangle$ using the commutator $[Z,X] = 2iY$ (Lecture~3, \S7.6):
\[
    \frac{d}{dt}\langle X\rangle = i\bigl\langle\tfrac{\Omega}{2}[Z,X]\bigr\rangle = i\cdot\tfrac{\Omega}{2}\cdot 2i\,\langle Y\rangle = -\Omega\,\langle Y\rangle.
\]
Similarly, using $[Z,Y] = -2iX$:
\[
    \frac{d}{dt}\langle Y\rangle = i\bigl\langle\tfrac{\Omega}{2}[Z,Y]\bigr\rangle = i\cdot\tfrac{\Omega}{2}\cdot(-2i)\langle X\rangle = +\Omega\,\langle X\rangle.
\]
And since $[Z,Z]=0$: $\frac{d}{dt}\langle Z\rangle = 0$.  Collecting:
\[
    \boxed{\frac{d}{dt}\begin{pmatrix}\langle X\rangle\\\langle Y\rangle\\\langle Z\rangle\end{pmatrix}
    = \Omega\begin{pmatrix}0&-1&0\\1&0&0\\0&0&0\end{pmatrix}
    \begin{pmatrix}\langle X\rangle\\\langle Y\rangle\\\langle Z\rangle\end{pmatrix}.}
\]
This is \textbf{Bloch precession}---the expectation-value vector rotates around the $z$-axis at angular frequency $\Omega$.  These are exactly the equations NMR engineers solve for nuclear spin precession in a static magnetic field.  The Ehrenfest equation has produced a classical equation of motion from a quantum system, using only linear algebra.

\medskip
\noindent In a future lecture on continuous systems, the same Ehrenfest equation applied to $H = p^2/2m + V(x)$ will yield $\frac{d}{dt}\langle x\rangle = \langle p\rangle/m$ and $\frac{d}{dt}\langle p\rangle = -\langle V'(x)\rangle$---literally Hamilton's equations for the expectation values.

\subsection{5.4 Energy conservation}
Set $L=H$:
\[
    \frac{d}{dt}\langle H \rangle = \frac{i}{\hbar}\langle [H,H] \rangle = 0,
\]
since any matrix commutes with itself.  Therefore the \textbf{average energy is conserved} for any time-independent system matrix.

\subsection{5.5 Time-translation invariance}
If $H$ does not depend explicitly on time, the system has \textbf{time-translation invariance}: the same experiment at a later time gives the same statistical results.  A time-dependent $H$ breaks this invariance and energy conservation.

% ============================================================
\section{6. Summary}
% ============================================================

\begin{enumerate}[nosep]
    \item The evolution ODE $i\hbar\,\dot{\mathbf{v}} = H\mathbf{v}$ is a first-order linear system with constant coefficients (an LTI system with skew-Hermitian coefficient matrix).
    \item Solution: $\mathbf{v}(t) = e^{-iHt/\hbar}\mathbf{v}(0)$.  The evolution operator $U(t)$ is unitary ($U^\dagger U = I$).
    \item Unitarity = norm preservation = information conservation (reversibility).
    \item Self-adjointness of $H$ follows from unitarity; $H$ is identified with energy.
    \item Energy eigenstates $H\mathbf{e}_k = E_k\mathbf{e}_k$ are stationary (normal modes).  General evolution: $\mathbf{v}(t) = \sum_k \alpha_k e^{-iE_k t/\hbar}\mathbf{e}_k$.
    \item The expectation value $\langle L \rangle = \mathbf{v}^\dagger L\mathbf{v}$ evolves via $\frac{d}{dt}\langle L \rangle = \frac{i}{\hbar}\langle [H,L] \rangle$.
    \item Average energy is automatically conserved: $\frac{d}{dt}\langle H \rangle = 0$.
\end{enumerate}

\end{document}
