\documentclass[12pt,a4paper]{article}
\usepackage[utf8]{inputenc}
\usepackage[T1]{fontenc}
\usepackage{amsmath,amssymb,amsthm}
\usepackage{geometry}
\usepackage{enumitem}
\usepackage{hyperref}
\usepackage{fancyhdr}
\usepackage{titlesec}

\geometry{margin=1in}
\pagestyle{fancy}
\fancyhf{}
\rhead{Lecture 2}
\lhead{Quantum Systems via Linear Algebra}
\cfoot{\thepage}

\titleformat{\section}{\large\bfseries}{}{0em}{}
\titleformat{\subsection}{\normalsize\bfseries}{}{0em}{}

\newtheorem{definition}{Definition}

\title{\textbf{Lecture 2: Eigenvectors of the Three Standard Observables} \\ \large States of a Two-Level System and Their Column-Vector Representations}
\author{Quantum Systems via Linear Algebra, Differential Equations, and Probability}
\date{}

\begin{document}
\maketitle
\thispagestyle{fancy}

% ============================================================
\section{1. Review: The Two-Level System}
% ============================================================

Recall from Lecture~1: a two-level system lives in $\mathbb{C}^2$ with standard basis vectors $\mathbf{e}_0$ and $\mathbf{e}_1$.  Three self-adjoint $2\times 2$ matrices $Z$, $X$, $Y$ serve as the standard observables, each with eigenvalues $+1$ and $-1$.

\medskip
\noindent\textbf{Key experimental facts:}
\begin{itemize}[nosep]
    \item No matter which observable is measured, the outcome is always $+1$ or $-1$---never any intermediate value.
    \item Measuring one observable randomizes the result of a different, non-commuting observable (see Lecture~1, Section~3.2).
    \item Repeating the \emph{same} measurement immediately yields the same result.
\end{itemize}

% ============================================================
\section{2. State Space and the Born Rule}
% ============================================================

The state space is $\mathbb{C}^2$ with the standard Hermitian inner product.  The eigenvectors of the observable $Z$ serve as the computational basis:
\[
    \mathbf{e}_0 = \begin{pmatrix}1\\0\end{pmatrix}\;(\text{eigenvalue }+1), \qquad
    \mathbf{e}_1 = \begin{pmatrix}0\\1\end{pmatrix}\;(\text{eigenvalue }-1).
\]
They are orthonormal: $\mathbf{e}_0^\dagger\mathbf{e}_1 = 0$, $\mathbf{e}_0^\dagger\mathbf{e}_0 = \mathbf{e}_1^\dagger\mathbf{e}_1 = 1$.

Any state can be written as a superposition:
\[
    \mathbf{v} = \alpha_0\,\mathbf{e}_0 + \alpha_1\,\mathbf{e}_1, \qquad
    \alpha_0,\alpha_1\in\mathbb{C}.
\]
By Axiom~3 (Born rule), measuring $Z$ on state $\mathbf{v}$ returns $+1$ with probability $|\alpha_0|^2$ and $-1$ with probability $|\alpha_1|^2$.  Total probability 1 requires $|\alpha_0|^2 + |\alpha_1|^2 = 1$, i.e.\ $\mathbf{v}^\dagger\mathbf{v}=1$.

\medskip
\noindent\textbf{Distinguishable states are orthogonal.}  Two states that can be unambiguously told apart by a single measurement are represented by orthogonal vectors.  $\mathbf{e}_0$ and $\mathbf{e}_1$ are distinguishable (measuring $Z$ tells them apart with certainty) and orthogonal.

\medskip
\noindent\textbf{Derivation: expectation value from the Born rule.}  Let $M = \sum_i \lambda_i\,\mathbf{e}_i\mathbf{e}_i^\dagger$ be the spectral decomposition.  Then
\[
    \langle M\rangle
    = \sum_i \lambda_i\,P(\lambda_i)
    = \sum_i \lambda_i\,|\mathbf{e}_i^\dagger\mathbf{v}|^2
    = \sum_i \lambda_i\,\mathbf{v}^\dagger\mathbf{e}_i\mathbf{e}_i^\dagger\mathbf{v}
    = \mathbf{v}^\dagger\!\Bigl(\sum_i \lambda_i\,\mathbf{e}_i\mathbf{e}_i^\dagger\Bigr)\mathbf{v}
    = \mathbf{v}^\dagger M\mathbf{v}.
\]
This simultaneously justifies the expectation-value formula and demonstrates the spectral theorem in action.

\medskip
\noindent\fbox{\parbox{\dimexpr\linewidth-2\fboxsep-2\fboxrule}{%
\textbf{Code says:} \texttt{psi.conj().T @ M @ psi} computes $\mathbf{v}^\dagger M\mathbf{v}$ directly; \texttt{sum(lam * abs(e.conj().T @ psi)**2 for lam, e in zip(evals, evecs.T))} computes $\sum_i \lambda_i|\mathbf{e}_i^\dagger\mathbf{v}|^2$.  Both return the same number.  \textit{(Verify with any state and any Pauli matrix.)}}}

% ============================================================
\section{3. Eigenvectors of the Three Standard Observables}
% ============================================================

Since $\mathbb{C}^2$ is two-dimensional, every state---including eigenvectors of $X$ and $Y$---must be a linear combination of $\mathbf{e}_0$ and $\mathbf{e}_1$.  We first write down the three observables as matrices, then derive their eigenvectors from symmetry constraints.

\subsection{3.0 The three standard observables as matrices}
The three standard observables for a two-level system are the \textbf{Pauli matrices}:
\[
    Z = \begin{pmatrix}1&0\\0&-1\end{pmatrix}, \qquad
    X = \begin{pmatrix}0&1\\1&0\end{pmatrix}, \qquad
    Y = \begin{pmatrix}0&-i\\i&0\end{pmatrix}.
\]
Each is Hermitian ($M = M^\dagger$) with eigenvalues $\pm 1$.  In Sections~3.1--3.3, we derive their eigenvectors from physical constraints; we then verify each result by direct matrix--vector multiplication.

\subsection{3.1 Eigenvectors of $X$}
A system in the $+1$ eigenvector of $X$, when measured with $Z$, gives $+1$ or $-1$ with probability $\tfrac{1}{2}$ each.  So
\[
    \mathbf{x}_+ = \alpha_0\,\mathbf{e}_0+\alpha_1\,\mathbf{e}_1, \qquad
    |\alpha_0|^2 = |\alpha_1|^2 = \tfrac{1}{2}.
\]
The simplest choice: $\alpha_0=\alpha_1=\frac{1}{\sqrt{2}}$, giving
\[
    \mathbf{x}_+ = \frac{1}{\sqrt{2}}\bigl(\mathbf{e}_0+\mathbf{e}_1\bigr) = \frac{1}{\sqrt{2}}\begin{pmatrix}1\\1\end{pmatrix}.
\]
The $-1$ eigenvector must be orthogonal to $\mathbf{x}_+$ and also satisfy $|\alpha_0|^2=|\alpha_1|^2=\tfrac{1}{2}$.  The unique (up to overall phase) solution:
\[
    \mathbf{x}_- = \frac{1}{\sqrt{2}}\bigl(\mathbf{e}_0-\mathbf{e}_1\bigr) = \frac{1}{\sqrt{2}}\begin{pmatrix}1\\-1\end{pmatrix}.
\]
\textbf{Check}: $\mathbf{x}_+^\dagger\mathbf{x}_-=\frac{1}{2}(1)(1)+\frac{1}{2}(1)(-1)=0$.\;$\checkmark$

\subsection{3.2 Reciprocal relationship}
We can invert to express $\mathbf{e}_0$ and $\mathbf{e}_1$ in terms of $\mathbf{x}_+$ and $\mathbf{x}_-$:
\[
    \mathbf{e}_0 = \frac{1}{\sqrt{2}}\bigl(\mathbf{x}_++\mathbf{x}_-\bigr), \qquad
    \mathbf{e}_1 = \frac{1}{\sqrt{2}}\bigl(\mathbf{x}_+-\mathbf{x}_-\bigr).
\]
The coefficients again have magnitude $\frac{1}{\sqrt{2}}$, confirming the symmetric relationship: starting from $\mathbf{x}_\pm$ and measuring $Z$ gives $\pm 1$ with equal probability.

\subsection{3.3 Eigenvectors of $Y$}
We need a third pair with the \emph{same} properties:
\begin{itemize}[nosep]
    \item $|\alpha_0|^2=|\alpha_1|^2=\tfrac{1}{2}$ (equal probability for $Z$-measurement).
    \item $\mathbf{y}_+^\dagger\mathbf{y}_-=0$ (distinguishable).
    \item When re-expressed in the $\{\mathbf{x}_+,\mathbf{x}_-\}$ basis, the coefficients must \emph{also} have magnitude $\frac{1}{\sqrt{2}}$ (equal probability for $X$-measurement).
\end{itemize}
These constraints uniquely determine (up to an overall phase):
\[
    \mathbf{y}_+ = \frac{1}{\sqrt{2}}\bigl(\mathbf{e}_0+i\,\mathbf{e}_1\bigr) = \frac{1}{\sqrt{2}}\begin{pmatrix}1\\i\end{pmatrix}, \qquad
    \mathbf{y}_- = \frac{1}{\sqrt{2}}\bigl(\mathbf{e}_0-i\,\mathbf{e}_1\bigr) = \frac{1}{\sqrt{2}}\begin{pmatrix}1\\-i\end{pmatrix}.
\]
The factor of $i$ is essential---real coefficients alone cannot produce a third symmetric pair.  To see why: any third pair $(\mathbf{w}_+, \mathbf{w}_-)$ with $\mathbf{w}_+ = \frac{1}{\sqrt{2}}(\mathbf{e}_0 + b\,\mathbf{e}_1)$ needs $|b|=1$ (for equal $Z$-measurement probabilities), $\mathbf{w}_+^\dagger\mathbf{w}_- = 0$ (distinguishability), \emph{and} $|\mathbf{x}_+^\dagger\mathbf{w}_+|^2 = \tfrac{1}{2}$ (equal $X$-measurement probabilities).  The last condition gives $|1+b|^2 = 2$, so $\operatorname{Re}(b)=0$.  With $|b|=1$ and $\operatorname{Re}(b)=0$ we get $b=\pm i$---a purely imaginary phase factor.  Real coefficients ($b\in\mathbb{R}$, $|b|=1$, $b=\pm 1$) only reproduce the $X$ eigenvectors.  Complex numbers provide the extra degree of freedom needed for the third basis.

\textbf{Check orthogonality}:
\[
    \mathbf{y}_+^\dagger\mathbf{y}_- = \tfrac{1}{2}(1)(1) + \tfrac{1}{2}(-i)(i) = \tfrac{1}{2}-\tfrac{1}{2}=0.\;\checkmark
\]

\textbf{Check in the $\{\mathbf{x}_+,\mathbf{x}_-\}$ basis}: substituting $\mathbf{e}_0=\frac{1}{\sqrt{2}}(\mathbf{x}_++\mathbf{x}_-)$ and $\mathbf{e}_1=\frac{1}{\sqrt{2}}(\mathbf{x}_+-\mathbf{x}_-)$:
\[
    \mathbf{y}_+ = \frac{1+i}{2}\,\mathbf{x}_+ + \frac{1-i}{2}\,\mathbf{x}_-.
\]
The magnitude of each coefficient: $\left|\frac{1\pm i}{2}\right|^2 = \frac{1+1}{4}=\frac{1}{2}$.\;$\checkmark$

\textbf{Matrix verification}: using the Pauli matrices from \S3.0,
\[
    Y\,\mathbf{y}_+ = \begin{pmatrix}0&-i\\i&0\end{pmatrix}\frac{1}{\sqrt{2}}\begin{pmatrix}1\\i\end{pmatrix}
    = \frac{1}{\sqrt{2}}\begin{pmatrix}(-i)(i)\\(i)(1)\end{pmatrix}
    = \frac{1}{\sqrt{2}}\begin{pmatrix}1\\i\end{pmatrix}
    = \mathbf{y}_+.\;\checkmark
\]

\subsection{3.4 Summary table}
\[
\renewcommand{\arraystretch}{1.4}
\begin{array}{c|c|c}
\textbf{Observable / eigenvalue} & \textbf{Expansion} & \textbf{Column vector} \\\hline
Z,\;+1 & \mathbf{e}_0 & \begin{pmatrix}1\\0\end{pmatrix} \\[4pt]
Z,\;-1 & \mathbf{e}_1 & \begin{pmatrix}0\\1\end{pmatrix} \\[4pt]
X,\;+1 & \dfrac{1}{\sqrt{2}}\bigl(\mathbf{e}_0+\mathbf{e}_1\bigr) & \dfrac{1}{\sqrt{2}}\begin{pmatrix}1\\1\end{pmatrix} \\[4pt]
X,\;-1 & \dfrac{1}{\sqrt{2}}\bigl(\mathbf{e}_0-\mathbf{e}_1\bigr) & \dfrac{1}{\sqrt{2}}\begin{pmatrix}1\\-1\end{pmatrix} \\[4pt]
Y,\;+1 & \dfrac{1}{\sqrt{2}}\bigl(\mathbf{e}_0+i\,\mathbf{e}_1\bigr) & \dfrac{1}{\sqrt{2}}\begin{pmatrix}1\\i\end{pmatrix} \\[4pt]
Y,\;-1 & \dfrac{1}{\sqrt{2}}\bigl(\mathbf{e}_0-i\,\mathbf{e}_1\bigr) & \dfrac{1}{\sqrt{2}}\begin{pmatrix}1\\-i\end{pmatrix}
\end{array}
\]

\subsection{3.5 Mutually unbiased bases}
Starting from any one of these six eigenvectors and measuring a \emph{different} observable always gives $P(+1)=P(-1)=\tfrac{1}{2}$.  The three pairs are symmetrically related: expanding any pair in terms of any other pair always produces coefficients of magnitude $\frac{1}{\sqrt{2}}$.

Three orthonormal bases with this pairwise property---every vector in one basis has equal-magnitude overlap $1/\sqrt{d}$ with every vector in every other basis---are called \textbf{mutually unbiased bases} (MUBs).  The $Z$, $X$, $Y$ eigenvectors form a \emph{maximal} set of three MUBs in $\mathbb{C}^2$, since the maximum number of MUBs in $\mathbb{C}^d$ is $d+1$ when $d$ is a prime power.  (The maximal number of MUBs in composite dimensions remains an open problem in quantum information theory.)

% ============================================================
\section{4. Counting Degrees of Freedom}
% ============================================================

A general state $(\alpha_0,\alpha_1)\in\mathbb{C}^2$ has \textbf{four real parameters} (real and imaginary parts of each component).  Two are physically irrelevant:
\begin{enumerate}[nosep]
    \item \textbf{Normalization}: $|\alpha_0|^2+|\alpha_1|^2=1$ removes one parameter.
    \item \textbf{Overall phase}: multiplying both components by $e^{i\theta}$ does not change any physical prediction.  Every measurable quantity involves $\mathbf{v}^\dagger M\mathbf{v}$; the common phase cancels.  This removes one more parameter.
\end{enumerate}
This leaves \textbf{two real parameters}.  These two parameters can be visualized as a point on the surface of a sphere---the \textbf{Bloch sphere}---which we construct explicitly in Lecture~3.

\medskip
\noindent\textbf{Aside: the true state space.}  Since both normalization and overall phase are physically irrelevant, the true state space is not $\mathbb{C}^2$ but the set of \emph{directions} in $\mathbb{C}^2$: the equivalence classes $\mathbf{v}\sim z\mathbf{v}$ for any $z\in\mathbb{C}^*$.  This quotient space is the complex projective line $\mathbb{CP}^1$, which is topologically the 2-sphere $S^2$---the Bloch sphere.  In intermediate calculations, non-unit vectors are fine; just renormalize via $\mathbf{v}\to\mathbf{v}/\sqrt{\mathbf{v}^\dagger\mathbf{v}}$ before extracting physics.

% ============================================================
\section{5. Summary}
% ============================================================

\begin{enumerate}[nosep]
    \item The state space of a two-level system is $\mathbb{C}^2$ with the standard Hermitian inner product.
    \item \textbf{Distinguishable} states are \textbf{orthogonal} vectors.
    \item The six eigenvectors of $Z$, $X$, $Y$ (three pairs) are determined, up to overall phase, by requiring equal probabilities ($\frac{1}{2}$) for measurements of the other two observables and mutual orthogonality within each pair.
    \item The eigenvectors of $Y$ require factors of $i$---real coefficients alone cannot produce a third symmetric pair.
    \item Only \textbf{two real parameters} characterize a physical state (after removing normalization and overall phase).
\end{enumerate}

\end{document}
